---
title: "Homework 4: Tidy data and joins"
output:
  github_document:
    toc: yes
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, collapse = TRUE, comment = "#>")
```

## Load packages

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(knitr))
```


# Data Reshaping (and relationship to aggregation)
**Problem**: I have data in one “shape” but wish it were in another. Usually this is because the alternative shape is superior for presenting a table, making a figure, or doing aggregation and statistical analysis.

## Functions

**Solution**: I reshape my data. For simple reshaping, `gather()` and `spread()` from tidyr will suffice. I could also try using `separate()` and `unite()`.

> `gather(data, key = "key", value = "value", ...)`: Collapse multiple columns into key-value pairs, duplicating all other columns as needed.

> `spread(data, key, value, ...)`: Spread a key-value pair across multiple columns.

> `separate(data, col, into, ...)`: Turns a single character column into multiple columns, given a regualr expression or a vector of character positions.

> `unite(data, col, ...)`: Convenience function to paste together multiple columns into one.


Resources:

* [tidyr in a Nutshell](https://github.com/trinker/tidyr_in_a_nutshell)
* [tidyverse documentation](https://tidyr.tidyverse.org/index.html)

## Activity 2: Life expectancy in Germany and New Zealand

For this activity I will make a tibble with one row per year and columns for life expectancy for Germany and New Zealand. I will challange myself by trying to write the least amount of code possible for this exercise and directly output a table and a scatterplot for comparison next to each other. I can leverage `spread()` to reshape my data for visual display and further scatterplot analysis. Let's start:



```{r, fig.align = 'top', fig.width = 12, fig.height = 7}
# select values for countries of interest and
# create the table using `spread()`
lifexp_table <- gapminder %>%
  filter(country %in% c("Germany", "New Zealand")) %>% 
  select(country, year, lifeExp) %>% 
  spread(key = country, value = lifeExp) # reshape data frame to multiple columns

# create the scatterplot using the new table
lifexp_plot <-  lifexp_table %>%
  ggplot(aes(`New Zealand`, `Germany`)) +
  geom_smooth(method = 'loess') +
  geom_point() +
  geom_text(aes(label=year),hjust=0, vjust=0) + # add year to point data
  labs(title="Life Expectancy in New Zealand and Germany",
       x="New Zealand (years)",
       y="Germany (years)")

grid.arrange(tableGrob(lifexp_table, rows=NULL, theme=ttheme_minimal()), lifexp_plot, nrow=1) # put table next to plot
```

We can of course also use `knitr::kable()` to make this table look pretty in my rendered homework. I however prefere the previously used `tableGrob` and use `grid.arrange` to display table and scatterpot next to each other.

```{r}
kable(lifexp_table)
```



# Join (join, merge, look up)
**Problem**: I have two data sources and need information from both in one new data object.

**Solution**: Perform a join, which borrows terminology from the database world, specifically SQL. 


**Functions**:

function name | purpose | type of join
-----------|---------------|----------
inner_join() | Returns all rows from x where there are matching values in y, and all columns from x and y. | mutating
semi_join() | Returns all rows from x where there are matching values in y, keeping just the columns from x. | filering
left_join() | Returns all rows from x, and all columns from x and y. | mutating
right_join() | Returns all rows from y, and all columns from x and y. | mutating
full_join() | Returns all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing. | mutating
anti_join() | Returns all rows from x where there are not matching values in y, keeping just columns from x. | filtering

In this activity I will create my own joins cheatsheet patterned after [Jenny's](http://stat545.com/bit001_dplyr-cheatsheet.html)

Other good reseources to read about joins are:

* The [Relational data chapter](http://r4ds.had.co.nz/relational-data.html) in R for Data Science.
* Handling [missing values](https://tidyr.tidyverse.org/reference/drop_na.html) in a dataframe.

### My data
I will work with two small data frames, `artists` and `songs`.

```{r}
artists <- tribble( # create a tibble
                    ~first, ~last, ~instrument, ~band,
                    "John", "Kay", "Vocals",  "Steppenwolf",
                    "George", "Harrison", "Guitar", "The Beatles",
                    "Mick", "Jagger", "Vocals", "Rolling Stones",
                    "Tom", "Jones", "Vocals", NA,
                    "Davy", "Jones", "Vocals", "",
                    "John", "Lennon", "Guitar", "The Beatles",
                    "Paul", "McCartney", "Bass", "The Beatles",
                    "Jimmy", "Page", "Guitar", "Led Zeppelin",
                    "Joe", "Perry", "Guitar", "Aerosmith",
                    "Keith", "Richards", "Guitar", "Rolling Stones", 
                    "Paul", "Simon", "Guitar", "Simon and Garfunkel",
                    "Ringo", "Starr", "Drums", "The Beatles",
                    "Joe", "Walsh", "Guitar", "The Eagles",
                    "Brian", "Wilson", "Vocals", "The Beach Boys",
                    "Nancy", "Wilson", "Vocals", "Heart",
                    "Wendy", "Davis", "Vocals", "Campfire Singer Songwriter"
                    )

songs <-   "
              song,               album,          writer,        band      
     Come Together,           Abbey Road,    John Lennon, The Beatles
          Dream On,            Aerosmith,   Steven Tyler,   Aerosmith
     Hello Goodbye, Magical Mystery Tour, Paul McCartney, The Beatles
  It's Not Unusual,     Along Came Jones,      Tom Jones,          "
  
songs <- read_csv(songs, skip = 1) 
# another way to create a tibble through loading a `.csv` file
```

### inner_join()

> `inner_join(x,y)` Return all rows from x where there are matching values in y, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned.

```{r}
inner_join <- inner_join(artists, songs)
kable(inner_join)
```

Inner joins only retain observations that appear in both datasets. Therefore we loose all band members that are not in either "The Beatles", "Aerosith" or have no value "NA". This also displays that `inner_join()` saves multiple matches. Here we automatically joined on the variable `band`, since our `songs` dataset displayes multiple of `The Beatles` songs. Every observation of a `song` from `The Beatles` is therefore connected to every artist (`first`, `last`) who played in `The Beatles` and we find the artist's name twice, connected to two `songs`. 


Using this join it is extremely important to be aware of "NA" values. I'll give an example:

```{r}
artists_NAvalues <- tribble( # create a tibble
                    ~first, ~last, ~instrument, ~band,
                    "Stefanie", "Lumnitz", "Guitar", NA,
                    "George", "Harrison", "Guitar", "The Beatles",
                    "Mick", "Jagger", "Vocals", "Rolling Stones",
                    "Tom", "Jones", "Vocals", NA)

inner_join(artists_NAvalues, songs)
```

Now suddenly me and Tom Jones wrote the song "It's Not Unusual" since the `inner_join()` used the variable `band` and me and Tom Jones are not in a band according to our dataset. Therefore if we work with dataframes in R it is always important to know your data and if necessary remove `NA` containing rows or replace the values to avoid miss information. We can use `drop_na()` for this:

```{r}
artists_clean <- drop_na(artists_NAvalues)
inner_join(artists_clean, songs)
```

### semi_join()

> `semi_join(x, y)`: Returns all rows from x where there are matching values in y, keeping just columns from x.

```{r}
semi_join(artists, songs) %>% 
  kable()
```

This is a really good example of how `semi_join()` can also be described as a filtering join. In comparison the the `inner_join()` the `semi_join()` will never duplicate matching rows of `artists` and therefore does not show us duplicated first and last names.


### left_join() 

> `left_join(x, y)`: Returns all rows from x, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned. This is a mutating join


Join matching rows from `b` to `a` by matching "x1" variable

```{r}
left_join(artists, songs) %>% 
  kable()
```

As we can see, this table includes all column names `first`, `last`, `instrument`, `band`,  `song`, `album`, `writer`. Futhermore all combinations of matches were returned as seen on the two rows for each name out of `The Beatles` for both songs in our data frame.

### right_join()


> `right_join()`: Returns all rows from y, and all columns from x and y. Rows in y with no match in x will have NA values in the new columns. If there are multiple matches between x and y, all combinations of the matches are returned.

```{r}
right_join(artists, songs) %>% 
  kable()
```

As we can see we lost our artists `Wendy`, due to the join on the `songs` dataframe this time. Otherwise the `right_join()` works like the `left_join()` only `x` and `y` play different roles. Following this logic, let's check if `left_join()` with reversed x and y contains the same information:


```{r}
left_join(songs, artists) %>% 
  kable()
```

And indeed, comparing the `first` column this join generated a dataframe containing the same information.


### full_join()

> `full_join()`: Returns all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing.

```{r}
full_join <- full_join(artists, songs)
kable(full_join)
```

Let's check if all rows are indeed kept:

```{r}
rows <- nrow(artists) + nrow(songs)
rows - nrow(full_join)
```

The comparison of rows and the visual display of all column names shows, all data from both datasets is kept, matching and non matching.


### anti_join():

> `anti_join()`: Returns all rows from x where there are not matching values in y, keeping just columns from x.

```{r, echo=FALSE, fig.height=7}
anti_join <- anti_join(artists, songs)

grid.arrange(tableGrob(anti_join, rows=NULL), tableGrob(songs, rows=NULL), ncol=1) # let's try another approach to visualizing tables
```

Additionally, I should mention that these joins all worked without specifyig a `by=`, due to one column name `bands` matching in both dataframes and the fucntion doing a natural join. If the column name is not matching but columns do contain identical information one can join by different variables on x and y use a named vector. For example, by = c("BANDS" = "band") will match x.BANDS to y.band.


