---
title: "Homework 4: Tidy data and joins"
output:
  github_document:
    toc: yes
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, collapse = TRUE, comment = "#>")
```

## Load packages

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(gridExtra))
```


# Data Reshaping (and relationship to aggregation)
**Problem**: I have data in one “shape” but wish it were in another. Usually this is because the alternative shape is superior for presenting a table, making a figure, or doing aggregation and statistical analysis.

## Functions

**Solution**: I reshape my data. For simple reshaping, `gather()` and `spread()` from tidyr will suffice. I could also try using `separate()` and `unite()`.

> `gather(data, key = "key", value = "value", ...)`: Collapse multiple columns into key-value pairs, duplicating all other columns as needed.

> `spread(data, key, value, ...)`: Spread a key-value pair across multiple columns.

> `separate(data, col, into, ...)`: Turns a single character column into multiple columns, given a regualr expression or a vector of character positions.

> `unite(data, col, ...)`: Convenience function to paste together multiple columns into one.


Resources:

* [tidyr in a Nutshell](https://github.com/trinker/tidyr_in_a_nutshell)
* [tidyverse documentation](https://tidyr.tidyverse.org/index.html)

## Activity 2: Life expectancy in Germany and New Zealand

For this activity I will make a tibble with one row per year and columns for life expectancy for Germany and New Zealand. I will challange myself by trying to write the least amount of code possible for this exercise and directly output a table and a scatterplot for comparison next to each other. I can leverage `spread()` to reshape my data for visual display and further scatterplot analysis. Let's start:



```{r, fig.align = 'top', fig.width = 12, fig.height = 7}
# select values for countries of interest and
# create the table using `spread()`
lifexp_table <- gapminder %>%
  filter(country %in% c("Germany", "New Zealand")) %>% 
  select(country, year, lifeExp) %>% 
  spread(key = country, value = lifeExp) # reshape data frame to multiple columns

# create the scatterplot using the new table
lifexp_plot <-  lifexp_table %>%
  ggplot(aes(`New Zealand`, `Germany`)) +
  geom_smooth(method = 'loess') +
  geom_point() +
  geom_text(aes(label=year),hjust=0, vjust=0) + # add year to point data
  labs(title="Life Expectancy in New Zealand and Germany",
       x="New Zealand (years)",
       y="Germany (years)")

grid.arrange(tableGrob(lifexp_table, rows=NULL, theme=ttheme_minimal()), lifexp_plot, nrow=1) # put table next to plot
```

We can of course also use `knitr::kable()` to make this table look pretty in my rendered homework. I however prefere the previously used `tableGrob` and use `grid.arrange` to display table and scatterpot next to each other.

```{r}
knitr::kable(lifexp_table)
```





# Join (join, merge, look up)
**Problem**: I have two data sources and need information from both in one new data object.

**Solution**: Perform a join, which borrows terminology from the database world, specifically SQL. 



In this activity I will create my own joins cheatsheet patterned after [Jenny's](http://stat545.com/bit001_dplyr-cheatsheet.html)

Other good reseources to read about joins are:
* The [Relational data chapter](http://r4ds.had.co.nz/relational-data.html) in R for Data Science.
* Handling [missing values](https://tidyr.tidyverse.org/reference/drop_na.html) in a dataframe.

### My data
I will work with two small data frames, `artists` and `songs`.

```{r}
artists <- tribble( # create a tibble
                    ~first, ~last, ~instrument, ~band,
                    "John", "Kay", "Vocals",  "Steppenwolf",
                    "George", "Harrison", "Guitar", "The Beatles",
                    "Mick", "Jagger", "Vocals", "Rolling Stones",
                    "Tom", "Jones", "Vocals", NA,
                    "Davy", "Jones", "Vocals", "",
                    "John", "Lennon", "Guitar", "The Beatles",
                    "Paul", "McCartney", "Bass", "The Beatles",
                    "Jimmy", "Page", "Guitar", "Led Zeppelin",
                    "Joe", "Perry", "Guitar", "Aerosmith",
                    "Keith", "Richards", "Guitar", "Rolling Stones", 
                    "Paul", "Simon", "Guitar", "Simon and Garfunkel",
                    "Ringo", "Starr", "Drums", "The Beatles",
                    "Joe", "Walsh", "Guitar", "The Eagles",
                    "Brian", "Wilson", "Vocals", "The Beach Boys",
                    "Nancy", "Wilson", "Vocals", "Heart",
                    "Wendy", "Davis", "Vocals", "Campfire Singer Songwriter"
                    )

songs <-   "
              song,               album,          writer,        band      
     Come Together,           Abbey Road,    John Lennon, The Beatles
          Dream On,            Aerosmith,   Steven Tyler,   Aerosmith
     Hello Goodbye, Magical Mystery Tour, Paul McCartney, The Beatles
  It's Not Unusual,     Along Came Jones,      Tom Jones,          "
  
songs <- read_csv(songs, skip = 1)
songs


```

### inner_join(artists, songs)

> `inner_join(x,y)` Return all rows from x where there are matching values in y, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned.

```{r}
inner_join <- inner_join(artists, songs)
```

Inner joins only retain observations that appear in both datasets. Therefore we loose all band members that are not in either "The Beatles", "Aerosith" or have no value "NA". This also displays that `inner_join()` saves multiple matches. Here we automatically joined on the variable `band`, since our `songs` dataset displayes multiple of `The Beatles` songs. Every observation of a `song` from `The Beatles` is therefore connected to every artist (`first`, `last`) who played in `The Beatles` and we find the artist's name twice, connected to two `songs`.

```{r load-xtable, echo = FALSE}
library(xtable)
```
<table border = 1>
<tr>
<td>
```{r results = 'asis', echo = FALSE}
print(xtable(artists, caption = "artists"),
      caption.placement = "top", include.rownames = FALSE, type = 'html')
```
</td>
<td>
```{r results = 'asis', echo = FALSE}
print(xtable(songs, caption = "songs"),
      caption.placement = "top", include.rownames = FALSE, type = 'html')
```
</td>
<td>
```{r results = 'asis', echo = FALSE}
print(xtable(inner_join, caption = "inner_join(artists, songs)"),
      caption.placement = "top", include.rownames = FALSE, type = 'html')
```
</td>
</tr>
</table>

Using this join it is extremely important to be aware of "NA" values. I'll give an example:

```{r}
artists_NAvalues <- tribble( # create a tibble
                    ~first, ~last, ~instrument, ~band,
                    "Stefanie", "Lumnitz", "Guitar", NA,
                    "George", "Harrison", "Guitar", "The Beatles",
                    "Mick", "Jagger", "Vocals", "Rolling Stones",
                    "Tom", "Jones", "Vocals", NA)

inner_join(artists_NAvalues, songs)
```

Now suddenly me and Tom Jones wrote the song "It's Not Unusual" since the `inner_join()` used the variable `band` and me and Tom Jones are not in a band according to our dataset. Therefore if we work with dataframes in R it is always important to know your data and if necessary remove `NA` containing rows or replace the values to avoid miss information. We can use `drop_na()` for this:

```{r}
artists_clean <- drop_na(artists_NAvalues)
inner_join(artists_clean, songs)
```


## Types of mutating join

### Let's join tibbles using four mutating functions:  `left_join`, `right_join`, `inner_join` and `full_join`.

### create two tibbles named `a` and `b`

```{r}
(a <- tibble(x1 = LETTERS[1:3], x2 = 1:3))
(b <- tibble(x1 = LETTERS[c(1,2,4)], x3 = c("T", "F", "T")))
```

### left_join: Join matching rows from `b` to `a` by matching "x1" variable

```{r}
left_join(a,b,by = "x1")
```

### right_join: Join matching rows from `a` to `b` by matching "x1" variable.

```{r}
right_join(a,b,by = "x1")
```

### inner_join: Join data. Retain only rows in both sets `a` to `b` by matching "x1" variable.

```{r}
inner_join(a, b, by="x1")
```

### full_join: Join data. Retain all values, all rows of `a` to `b` by matching "x1"

```{r}
full_join(a,b, by="x1")
```

### what happen if we do not specify `by` option?

```{r}
left_join(a,b)
```

### what happen if we join two different variables (e.g., "x1" to "x3") from two tibbles `a` to `b`?

```{r}
left_join(a,b, by= c("x1" = "x3"))
```

### what happen if two columns of `a` and `c` datasets have the identical colnames?

```{r}
# make data frame c and use inner_join()
(c <- tibble(x1 = c(LETTERS[1:2],"x"), x2 = c(1,4,5)))
left_join(a,c)
```

## In class practice

`nycflights13` dataset has several tibbles e.g., `flights`, `airports`, `planes`, `weather`.

### 1. Explore `nycflights13` dataset

```{r}
#check the tibbles included in `nycflights13` package
class(flights)

colnames(flights)

colnames(airlines)

colnames(weather)
```

### 2. Drop unimportant variables so it's easier to understand the join results. Also take first 1000 rows to run it faster.

```{r}
flights2 <- flights[1:1000,] %>% 
  select(year, tailnum, carrier, time_hour)
flights2
dim(flights2)
```

### 3. Add airline names to `flights2` from `airlines` dataset.

```{r}
# Which join function to use?
colnames(airlines)
colnames(flights2)
left_join(flights2, airlines)
```

### 4. Add `weather` information to the `flights2` dataset by matching "year" and "time_hour" variables.

```{r}
colnames(weather)
flights2 %>%
  left_join(weather, by = c("year", "time_hour"))
```

### 5. Add `weather` information to the `flights2` dataset by matching only "time_hour" variable

```{r}
flights2 %>%
  left_join(weather, by = c("time_hour"))
```

## Types of filtering join

### Let's filter tibbles using two filtering functions: `semi_join`, `anti_join` 


### example for `semi_join`: All rows in `a` that have a match in `b`

```{r}
semi_join(a, b, by = "x1")
```

### example for `anti_join`: All rows in `a` that do not have a match in `b`

```{r}
anti_join(a, b)
```

### example of joinin by matching two variables (e.g., "x1", "x2") from both datasets `a` and `c`

```{r}
a
c
semi_join(a, c)
```

## Types of Set Operations for two datasets

### Let's use three `set` functions: `intersect`, `union` and `setdiff` 

### create two tibbles named `y` and `z`, similar to Data Wrangling Cheatsheet

```{r}
(y <-  tibble(x1 = LETTERS[1:3], x2 = 1:3))
(z <- tibble(x1 = c("B", "C", "D"), x2 = 2:4))
```

### example for `intersect`: Rows that appear in both `y` and `z`

```{r}
intersect(y, z)
```

### example for `union`: Rows that appear in either or both `y` and `z`

```{r}
union(y,z)
```

### example for `setdiff`: Rows that appear in `y` but not `z`. __Caution:__ `setdiff` for `y` to `z` and `z` to `y` are different.

```{r}
setdiff(y, z)
setdiff(z, y)
```

### what happen if colnames are differentin `y` and `x`? Is there any error message and why?

```{r}
(x <- tibble(x1 = c("B", "C", "D"), x3 = 2:4))
#intersect(y,x)
x
y
```

## Types of binding datasets

### Let's bind datasets by rows or column using two binding functions:

### example for `bind_rows`: Append `z` to `y` as new rows

```{r}
bind_rows(y,z)
```

### example for `bind_cols`: Append `z` to `y` as new columns. __Caution__: matches rows by position. Check colnames after binding.

```{r}
bind_cols(y,z)
```

### what happen if colnames are different between `y` and `x` datasets?

```{r}
bind_cols(y,x)
```

##Practice Exercises
Practice these concepts in the following exercises. It might help you to first identify the type of function you are applying.


### 1. Filter the rows of `flights2` by matching "year" and "time_hour" variables to `weather` dataset. Use both `semi_join()` and `anti_join()`

```{r}
semi_join(flights2, weather)
anti_join(flights2, weather)
```

### 2. Can we apply `set` and `binding` funcions between `flights2` and `weather` datasets. Why and why not?

```{r}
# union(flights2, weather)
bind_rows(flights2,weather)
# bind_cols(flights2, weather)
```

### 3. Let's create a tibble `p` with "x1" and "x2" coulmns and have duplicated element in "x1" column. Create another tibble `q` with "x1" and "x3" columns. Then apply `left_join` function `p` to `q` and `q` to `p`. 

```{r}
(p <- tibble(x1 = c("a", "b", "b", "c"), x2 = 1:4))
(q <- tibble(x1 = c("a", "b", "c", "c"), x3 = 5:8))
left_join(p,q)
left_join(q,p)
```